---
title: "Excercise Classification based on Activity Dataset"
output: html_document
bibliography: references.bib
---

## Summmary

This is a study to analyze if you can predict exercise based on fitness readings. The data sets contain information measured to establish and categorize exercise activity based on various sensor data [@author2013].

Predictions became quite accurate using a bagged tree method when explicitly looking at the data which was present in both the training and testing sets
while using principal component analysis to reduce the dimensionality. Due to this, the normal visual analysis is not present. There is an accuracy table shown.
When doing validation, the accuracy is 99.95%.

```{r}
# import the required libraries
library(dplyr)
library(caret)
library(ISLR)
library(rattle)
```

```{r}
# Get the training and testing datasets from online and save to variables
training <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
testing <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
```
## Clean the data set before training
```{r}
# Remove near-zero variance predictors
nzv <- nearZeroVar(training)
training <- training[, -nzv]

# Select relevant columns
training <- training[, 8:ncol(training)]
testing <- testing[, 8:ncol(testing)]

# Remove the problem_id column from testing
testing <- testing %>% select(-problem_id)

# Remove columns with all NA values
training <- training %>% select(where(~ !all(is.na(.))))
testing <- testing %>% select(where(~ !all(is.na(.))))

# Convert columns to numeric, except the last column in training which is 'classe'
training <- training %>%
        mutate(across(
                -ncol(training),
                as.numeric
        ))

# Ensure 'classe' is a factor in training
training$classe <- factor(training$classe)

# Identify common columns between training and testing datasets
common_cols <- intersect(names(training), names(testing))

# Select common columns from both datasets
training <- training[, c(common_cols, "classe")]

# Split training for actual training and validation
inTrain <- createDataPartition(y = training$classe, p = 0.9, list = FALSE)
validation <- training[-inTrain, ]
training <- training[inTrain, ]

testing <- testing[, common_cols]
```

## Train the model
```{r cache=TRUE}
set.seed(42069)
# Set seed for reproducibility
set.seed(42069)

# Train the model
modelFit <- train(classe ~ ., data = training, method = "treebag", preProcess = c("pca"), na.action = na.pass)
```

## Model details and statistics
```{r}
# Summary of the final model
compact_summary <- function(model) {
        cat("Model Method:", model$method, "\n")
        cat("Number of Trees:", length(model$finalModel$frame), "\n")
        cat("Model Accuracy:", max(model$results$Accuracy), "\n")
        cat("Model Parameters:\n")
        print(model$bestTune)
}
compact_summary(modelFit)

confusionMatrix(predict(modelFit, newdata = validation), validation$classe)
```

## Get the results from the testing dataset
```{r}
# Predict on the testing dataset
pred <- predict(modelFit, newdata = testing)
print(pred)
```

